---
title: "Vector PAR Model Summary - Simulation"
author: 'Group 3: Abby Foes, Zheng Lan, Justin Landis, Yu Liu, Alec Reinhardt'
date: "`r Sys.Date()`"
output: html_document
---
```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(tibble)
library(ggplot2)
```

$$
y_{it} \sim \text{Poisson}(m_{t})
$$
$$
m_t = \sum_{l=1}^{q} \beta_{l} y_{t-l} +
\left(1 - \sum_{l=1}^{q} \beta_{l} \right) \cdot \exp(\mathbf{x}_t^\top \boldsymbol{\gamma})
$$
Where:

* $\beta_{l}$: AR coefficients (constrained so sum $\leq$ 1)
* $\gamma$: covariate effect vector
* $\tau$: mixing weight for AR component

Priors:

* $\boldsymbol{\gamma} \sim \mathcal{N}(\boldsymbol{\mu}, \Sigma_\gamma)$
* $\Sigma_\gamma \sim \text{Inv-Wishart}(\nu, \Psi)$ — (approximated as fixed in Stan)
* $\tilde{\boldsymbol{\beta}} \sim \text{Dirichlet}(\alpha)$, $\beta = \tau \tilde{\boldsymbol{\beta}}$
* $\tau \sim \text{Beta}(a, b)$

We fit this model **independently for each item**.  

# Simulation Parameters
```{r}
devtools::load_all()
set.seed(2025)
sim_data <- simulate_vector_par(n_items = 12, n_brands = 3, T = 100, q = 1)
true_mu_brand <- attr(sim_data, "true_params")$mu_brand
true_eta <- attr(sim_data, "true_params")$eta
true_f <- attr(sim_data, "true_params")$f
```

# Prepare Stan Data and Fit
# Prepare Stan Data and Fit (One Item Example)
```{r}
item1 <- sim_data %>% filter(item_id == 1)
fit <- fit_par_item(item1, q = 1)
```

# Posterior Summaries
```{r}
summary_gamma <- fit$summary(variables = paste0("gamma[", 1, "]"))
summary_tau <- fit$summary(variables = "tau")
summary_beta <- fit$summary(variables = paste0("beta[", 1, "]"))
```

# True vs Estimated (Example)
```{r}
true_gamma1 <- attr(sim_data, "true_params")$gamma[[1]][1]

tibble(
  parameter = c("gamma[1]", "tau", "beta[1]"),
  estimate = c(summary_gamma$mean, summary_tau$mean, summary_beta$mean),
  lower = c(summary_gamma$q5, summary_tau$q5, summary_beta$q5),
  upper = c(summary_gamma$q95, summary_tau$q95, summary_beta$q95),
  true = c(true_gamma1, NA, NA)
)
```

# Full loop for independent item fitting
```{r}
# Get item-level data splits
item_list <- split(sim_data, sim_data$item_id)

# Loop and fit each item independently
fits <- purrr::map(item_list, ~ fit_par_item(.x, q = 1))
names(fits) <- paste0("item_", names(item_list))

# Helper to pull posterior mean of gamma[1], tau, and beta[1] for each item
extract_summary <- function(fit) {
  gamma_mean <- fit$summary(variables = "gamma[1]")$mean
  tau_mean <- fit$summary(variables = "tau")$mean
  beta_mean <- fit$summary(variables = "beta[1]")$mean
  tibble(gamma = gamma_mean, tau = tau_mean, beta = beta_mean)
}

summary_df <- purrr::map_dfr(fits, extract_summary, .id = "item_id")
summary_df$item_id <- as.integer(gsub("item_", "", summary_df$item_id))

# Add true gamma for comparison
true_gammas <- purrr::map_dbl(attr(sim_data, "true_params")$gamma, 1)
summary_df$true_gamma <- true_gammas[summary_df$item_id]

summary_df

ggplot(summary_df, aes(x = true_gamma, y = gamma)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  labs(title = "Posterior Mean of γ vs True γ (per item)", x = "True γ", y = "Estimated γ") +
  theme_minimal()
```



```{r}

brands <- c("B1", "B2", "B3", "B4")
brand_data <- lapply(brands, function(b) filter(data_set_tidy, brand == b))
names(brand_data) <- brands

# brand_fits <- lapply(brand_data, function(df) {
#   item_list <- split(df, df$item)
#   item_list <- split(df, df$item)
# 
#   # Ensure correct column names for fitting
#   item_list <- purrr::map(item_list, ~ .x %>%
#     rename(y = QTY) %>%
#     mutate(time = row_number()))  # ensure time column exists
#   map(item_list, ~ fit_par_item(.x, q = 1))
# })

```



